---
title: "(week 2 lecture 1) tidyverse"
author: "Divya Shridar"
date: "10/19/2022"
output: html_document
---

## Package Installation

```{r Packages}
library(tidyverse)
library(dplyr)
```

## Setup Workplace
```{r}
# path <- dirname(rstudioapi::getSourceEditorContext()$path) # get path where this current r markdown is located
# input_dir <- path
# setwd(path) # set working directory to current path
```

## Import Data as Tibble

```{r Data Import}
LP_full <- read.csv('London_Paris_full.csv',
                      header = TRUE)
head(LP_full)

geno_data <- read.csv('London_Paris_Jan_geno.csv', 
                      header = TRUE)
head(geno_data)
```

## Save as Tibble

```{r As Tibble}
LP_Jan_geno <- as_tibble(geno_data)
```
## Explore Dataset

```{r}
dim(LP_Jan_geno) # has 500 rows and 14 columns
str(LP_Jan_geno) # columns include factor, doubles, integers and a logical; NAs present as well.
sum(is.na(LP_Jan_geno$X2)) # 500 of 500 is NA

  # to note:
    # date not coded in required format of YYYYMMDD
    # contains SNP data coded as integers
    # there is a column X2 that seeems to only contain NA
```

## Begin Steps 

### Obtain the relevant characteristics from the London_Paris_full data for the genotyped samples contained in London_Paris_Jan_geno
```{r}
# we need specific columns of "sample_id","dob","dor","dod","dodiag","cc_status","smoke_status","vit_status","a_sta_smok","a_quit_smok","age_recr"

required_colnames<-c("sample_id",
                     "dob",
                     "dor",
                     "dod",
                     "dodiag",
                     "cc_status",
                     "smoke_status",
                     "vit_status",
                     "a_sta_smok",
                     "a_quit_smok",
                     "age_recr")
#length(required_colnames) #11; just to check

# we also want all rows of LP_Jan_geno aka for each sample id
required_samples<-LP_Jan_geno$sample_id
#length(required_samples) # 500; just to check

# filter to get relevant information in new object
  # via filter(data, ) from LP_full for rows
  # via select for specific columns for columns
London_Paris_Jan_info <- filter(LP_full,
                                 sample_id %in% required_samples) %>% 
  select(required_colnames)
head(London_Paris_Jan_info) # 500 rows and 11 cols
```

### Merge the genotype information to the data obtained in the previous step, making sure that we don’t mix up the sample_ids.
```{r}
# prepare info to be merged
London_Paris_Jan_geno <- LP_Jan_geno %>% select(sample_id,  # all the columns we want to keep
                                                starts_with("SNP")) 
London_Paris_Jan_geno
London_Paris_Jan_info

# merge via merge()
  # alternative is left join!
  # LP_Jan_info_geno <- left_join(London_Paris_Jan_info, London_Paris_Jan_geno, by="sample_id")
LP_Jan_info_geno <- merge(London_Paris_Jan_info,
                          London_Paris_Jan_geno,
                          by = "sample_id") # make sure we join on appropriate column aka the sample_id is common across both datasets

LP_Jan_info_geno # 500 rows, 16 columns
```

### Transform relevant columns into factors and dates
```{r}
# transform via mutate()
LP_Jan_info_geno <- LP_Jan_info_geno %>% mutate(
  sample_id = as.character(sample_id), # was already character tho
  dob = as.Date(dob,"%d%b%Y"), # change character to date via as.Date
  dor = as.Date(dor,"%d%b%Y"),
  dod = as.Date(dod,"%d%b%Y"),
  dodiag = as.Date(dodiag,"%d%b%Y"),
  cc_status = as.factor(cc_status), # change binary integer (0/1) to factor
  smoke_status = as.factor(smoke_status),
  vit_status = as.factor(vit_status)
)

LP_Jan_info_geno
```
### Recode all dates so that they are in the format YYYYMMDD
```{r}
# use mutate again to apply a function
  # first write function so that can be applied to all rows
    # want to change 1928-12-27	to YYYYMMDD aka remove hyphens
sub_hifen <- function(x){
  x <- str_replace_all(x, "-", "") # applies to single row
    # str_replace_all better than str_replace that only replaces first -
  #print(x) #just for my checking
}

  # second, mutate so apply to all rows
LP_Jan_info_geno <- LP_Jan_info_geno %>%
  mutate_at(vars(starts_with("d")), # mutate at columns that start with d via vars()
            sub_hifen) # function to be applied

LP_Jan_info_geno
```

### Make sure 120 and 19000101 are coded when required
```{r}
# dod should be coded as 19000101 for alive participants;
  # use if else
# dodiag should be coded as 19000101 for controls;
# a_sta_smok should be coded as 120 for never smokers;
# a_quit_smok should be coded as 120 for current and never smokers.

LP_Jan_info_geno <- LP_Jan_info_geno %>% 
  mutate(dod = ifelse(vit_status == 1 & is.na(dod), # alive participants
                      19000101, # change if yes
                      dod), # keep current val if no
         dodiag = ifelse(cc_status == 0 & is.na(dodiag),
                         19000101,
                         dodiag),
         a_sta_smok = ifelse(smoke_status==1 & is.na(a_sta_smok),
                             120,
                             a_sta_smok),
         a_quit_smok = ifelse(is.na(a_quit_smok) & (smoke_status == 1 | smoke_status == 3),
                              120,
                              a_quit_smok))
LP_Jan_info_geno
```

### Confirm that the levels of the data are as expected. For example, there is no other level then 3 in the smoke_status column or that there are only two levels in the vit_status column
```{r}
# Recall that the levels should be as follows:
  # cc_status, 0-controls, 1-cases
  # smoke_status, 1-never smokers, 2-former smokers, 3-current smokers
  # vit_status, 1-alive, 2-dead
  # SNP colums should only have 3 levels, 0,1 and 2.

# check via lapply(X = , FUN = )
inspect_levels <- select(LP_Jan_info_geno, # df to select from 
                         cc_status, # required columns to check
                         smoke_status, # required columns to check
                         vit_status, # required columns to check
                         starts_with("SNP"))  %>%  
  lapply(table) # get a table summary of counts of the values for each column to check for
inspect_levels

# apply() - applies a function over an array of margins
# lapply() - applies a function over a list or a vector
# sapply() - is a friendlier version of lapply and allows us to return, when appropriate, a vector or a matrix. In the example above we could have used sapply as well.

  # Essentially, the choice of the _apply() function will depend on what is your input and what you want as an output.
  # In this case we wanted to return a list where each element of the list contains a table with the counts for each column of interest. 
  # lapply() returns a list. 
  # We can use also use the sapply function and the apply() functions here as well and will obtain the same result.
```

```{r}
# From what we can see the columns smoke_status and vit_status have some unwanted categories. 
  # We will remove these from our data by filtering them out
LP_Jan_info_geno <- LP_Jan_info_geno %>%
                        filter(smoke_status!=4, # remove where =4
                               vit_status %in% c(1,2)) # keep only where 1 and 2

dim(LP_Jan_info_geno) # 485 rows,  16 columns

# check filtered columns
table(LP_Jan_info_geno$smoke_status)
table(LP_Jan_info_geno$vit_status)
```

### Check if there are any NAs remaining
```{r}
# sum NA present in each column
NA_test <- select(LP_Jan_info_geno,
                  sample_id:age_recr) %>% 
  lapply(is.na) %>% # gives FALSE/TRUE for every row in every column chosen above
  sapply(sum,
         simplify = TRUE) # more consolidated table instead of single row
NA_test # NA present in a_sta_smok and a_quit_smok

# remove NA present via filter()
LP_Jan_info_geno <- filter(LP_Jan_info_geno,
                           !is.na(a_sta_smok), # true for is not NA
                           !is.na(a_quit_smok))
dim(LP_Jan_info_geno) #326 rows, 16 columns
```

### Arranging our data by cc_status and sample_id so that controls appear first.
```{r}
# arrange rows by columns via arrange()
LP_Jan_info_geno
LP_Jan_info_geno <- arrange(LP_Jan_info_geno,
                            cc_status, # rows ascend from this first
                            sample_id) # within each cc_status category, ascending via sample_id
LP_Jan_info_geno
```

### Save our final object as an .rds object.
```{r}
# save as an r data file
saveRDS(LP_Jan_info_geno,
        "London_Paris_info_geno.rds")
```

## Challenge
```{r}
LP_Jan_info_geno
London_Paris_info_geno_checked <- filter(LP_Jan_info_geno,
       dodiag>dor & ### the date of diagnosis has to be greater than the recruitment date
         dodiag>dob & ### the date of diagnosis has to be greater than the birth date
         dor>dob & ### the recruitment date has to be greater than the birth date
         dod>dor &  ### the date of death has to be greater than the recruitment date
         ifelse(cc_status==1 & vit_status==2, ### if a patient has been diagnosed with lung cancer and the patient died, than the death date has to be greater than the diagnosis date
                dod>dodiag,
                TRUE) &
       ifelse(vit_status==1,# if the vital_status = 1 then no death date should be observed
              dod==19000101,
              TRUE) &
       ifelse(vit_status==2,# if the vital_status = 2 then a death date should be observed
              dod!=19000101,
              TRUE) &
       ifelse(dod!=19000101, #if there is a death date then the death date should be greater than the birth date
              dod>dob,
              TRUE) &
         ifelse(cc_status==1, # if an observation is a case then there must be a diagnosis date
                is.na(dodiag)!=TRUE,
                TRUE) &
         ifelse(cc_status==0, # if an observations is not a case then there shouldn’t be a diagnosis date
                is.na(dodiag)==TRUE,
                TRUE) &
         ifelse(smoke_status==2 | smoke_status==3,
               a_sta_smok!=120,
               TRUE) &
         ifelse(smoke_status==2,
                a_quit_smok!=120,
                TRUE) &
         ifelse(smoke_status==2,
                a_quit_smok>=a_sta_smok,
                TRUE)) 
```

```{r}
dim(London_Paris_info_geno_checked) # 48 rows, 16 columns?
London_Paris_info_geno_checked
```

```{r}
# save
saveRDS(London_Paris_info_geno_checked,
        "London_Paris_info_geno_checked.rds")
```

